---
title: 'Lab 3 : ggplot'
author: "Jeremy Morris"
date: "2/20/2019"
output: 
  html_document: 
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(here)
```


# Introduction to ggplot2

“ggplot()” is the generic function call. A plot can include any of the following objects.

* `ggplot` - the primary building block of a chart. By itself only generates a blank chart.
* `aes` - short for 'aesthetic mapping'. This function tells the plot how to map data to different portions of the chart (e.g. x-axis, y-axis, fill, etc.)
* `geom` - short for 'geometric object'. These are the actual charts, a short list of options follows.
    * `geom_point` - scatterplot
    * `geom_bar` - bar chart
    * `geom_line` - line chart
* `labs` - chart labels and titles
* `coord` - functions to modify the coordinate space. The most commonly used is `coord_flip` to rotate a chart 90 degrees.
* `facet` - these control how many charts are shown. We will cover these in more detail in the next lab.
* `theme` - functions to modify the overall look and feel over the chart.

The great thing about ggplot is that it allows you to control nearly every aspect of a data visualization. Because there is so much control, there is also a learning curve associated with that level of control.

The basic structure of a ggplot call is to first call the ggplot function and then add the different components to the chart. The aesthetic mappings can be added when `ggplot` is called or they can be added in the individual geometric objects. If added in the `ggplot` call, these aesthetic mappings will be used in all `geom` functions. You would add them to the `geom` function if you wanted to override the settings in `ggplot`.


### mtcars | weight vs. miles per gallon

In this example, we have a basic `ggplot` call with data and aesthetic mappings in the `ggplot` function.
```{r scatter}
ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point()
```

Here we produce the same chart but specify aesthetic mappings along with the `geom`. Note that the two charts are exactly the same.
```{r scatter2}
ggplot(mtcars) +
    geom_point(aes(x = wt, y = mpg))
```

If you want to get very specific, the data used can also be added at the `geom` level. This allows you to build a visualization on multiple data sets (if that's something you want to do). Going forward we will likely specify data and aesthetic mappings in the `ggplot` function but wanted to point out that there is some flexibility with how/where you specify those things.

```{r scatter3}
ggplot() +
    geom_point(aes(x = wt, y = mpg),data=mtcars)
```

### heightweight | height vs. age  
* df.heightweight is an object of type tibble
* it has 236 observations and 5 variables
* variables can be accessed by the convention “df.heightweight$ageYear”
* variables have different types themselves, which costrain how we can visualize them effectively

```{r hw,message=FALSE}
df.heightweight <- read_csv(here('_data','heightweight.csv'))
df.heightweight

ggplot(df.heightweight, aes(x = ageYear, y = heightIn)) + 
  geom_point()
```

* In summary: Our objective is similar to that in the previous example.  However, we are using a different dataset (heightweight) and are interested in the relationship between height and age.

### heightweight | height vs. age | include average height as reference 

Add the average height to the plot for the reader's reference. Note that we're adding a second `geom` to the chart and even overriding the aesthetic mapping. This is a good example of how to add extra/aggregated data to a chart.
```{r hw_scatter2}
ggplot(df.heightweight, aes(x = ageYear, y = heightIn)) +
  geom_point() +
  geom_hline(yintercept=mean(df.heightweight$heightIn))
```

### heightweight | height vs. age by gender
In this example, we add the 'color' aesthetic to the chart. You can see that `ggplot` automatically recognizes the different levels in the 'sex' variable and colors them differently.

```{r hw3}
ggplot(df.heightweight, aes(x = ageYear, y = heightIn, colour = sex)) +  
  geom_point()
```

If we want a simple bar chart showing the average height by sex, we first pre-process the data calculating the average we want and then use `geom_col`.

```{r hw_col}
by_sex <- df.heightweight %>% group_by(sex) %>% summarise(mean_height=mean(heightIn))
ggplot(by_sex) +
  geom_col(aes(x=sex,y=mean_height))
```

Not a super helpful chart, but it illustrates how to make a bar chart. We could also look at a boxplot for the same data to look at the distribution of values. In this example, we dress up the chart a little with some color and an added theme to make it look a little cleaner.

```{r hw_col2}
ggplot(df.heightweight) +
  geom_boxplot(aes(x=sex,y=heightIn,fill=sex)) +
  theme_bw() +
  labs(x='Sex',y='Height (in)') +
  theme(legend.position='none')
```

We can also look at a density plot for each using the fill aesthetic and alpha blending to make a good looking chart. Note that we leave the legend on this one because `sex` doesn't show up on an axis.

```{r hw_density}
ggplot(df.heightweight) +
  geom_density(aes(x=heightIn,fill=sex),alpha=0.6) +
  labs(x='Height (in)',y='Density',fill='Sex') +
  theme_bw()
```

### heightweight | height vs. age by weight
In this example, we see a continuous variable specified for the color aesthetic and note that `ggplot` uses a continuous color scale.
```{r hw4}
ggplot(df.heightweight, aes(x = ageYear, y = heightIn, colour = weightLb)) +
  geom_point()
```


Here, we modify the size of the points using a static value.
```{r hw5}
ggplot(df.heightweight, aes(x = ageYear, y = heightIn, colour = weightLb)) +
  geom_point(size = 5)
```

Now the points remain the same size but we add transparency.
```{r hw6}
ggplot(df.heightweight, aes(x = ageYear, y = heightIn, colour = weightLb)) +
  geom_point(size = 5, alpha = 0.5)
```

***

### Create line graph from text - step by step
```{r txt_read,message=FALSE}
df_line_graph <- read_csv(here('_data','line.csv'))
df_line_graph
```

basic line graph
```{r txt_line}
ggplot(df_line_graph, aes(x=monthNum, y = volume, colour=type)) + 
  geom_line()
```

remove gridelines
```{r txt_remove}
ggplot(df_line_graph, aes(x=monthNum, y = volume, colour=type)) +
  geom_line() +
  theme_classic()
```

add points to follow example from text
```{r}
ggplot(df_line_graph, aes(x = monthNum, y = volume, colour = type)) +
  geom_line() +
  theme_classic() +
  geom_point(size = 4)
```

fix the x-axis
```{r}
ordered.month.list <- unique(as.character(df_line_graph$month))
df_line_graph$month <- factor(df_line_graph$month, levels=ordered.month.list)
ggplot(df_line_graph, aes(x = month, y = volume, colour = type, group = type)) +
  geom_line() +
  theme_classic()
```

label lines directly
```{r}
ggplot(df_line_graph, aes(x=month, y = volume, colour = type, group = type)) +
  geom_line() +
  theme_classic() +
  geom_text(data = df_line_graph[df_line_graph$month == "Dec", ], aes(label = type)
            , hjust=0.9, vjust=-0.5) +
  theme(legend.position=("none"))
```

change the colors of the series
```{r}
ggplot(df_line_graph, aes(x=month, y = volume, colour = type, group = type)) +
  geom_line() +
  theme_classic() +
  geom_text(data = df_line_graph[df_line_graph$month == "Dec", ], aes(label = type),
            hjust=0.9, vjust=-0.5) +
  scale_color_manual(values= c("#3182bd", "#bdbdbd")) +
  theme(legend.position=("none"))
``` 

increase the thickness of series lines
```{r}
ggplot(df_line_graph, aes(x=month, y = volume, colour = type, group = type)) +
  geom_line(size=1) +
  theme_classic() +
  geom_text(data = df_line_graph[df_line_graph$month == "Dec", ], aes(label = type)
            , hjust=0.9, vjust=-0.5) +
  scale_color_manual(values= c("#3182bd", "#bdbdbd")) +
  theme(legend.position=("none"))
```

change the color of the axis lines to grey
```{r}
ggplot(df_line_graph, aes(x=month, y = volume, colour = type, group = type)) +
  geom_line(size=1) +
  theme_classic() +
  geom_text(data = df_line_graph[df_line_graph$month == "Dec", ], aes(label = type), 
            hjust=0.9, vjust=-0.5) +
  scale_color_manual(values= c("#3182bd", "#bdbdbd")) +
  theme(legend.position=("none"),axis.line = element_line(color = "#bdbdbd"))
```

change the color of the axis ticks to grey
```{r}
ggplot(df_line_graph, aes(x=month, y = volume, colour = type, group = type)) +
  geom_line(size=1) +
  theme_classic() +
  geom_text(data = df_line_graph[df_line_graph$month == "Dec", ], aes(label = type),
            hjust=0.9, vjust=-0.5) +
  scale_color_manual(values= c("#3182bd", "#bdbdbd")) +
  theme(legend.position=("none"),
        axis.line = element_line(color = "#bdbdbd"),
        axis.ticks = element_line(color = "#bdbdbd")
  )
```

change the color of the axis labels to grey
```{r}
ggplot(df_line_graph, aes(x=month, y = volume, colour = type, group = type)) +
  geom_line(size=1) +
  theme_classic() +
  geom_text(data = df_line_graph[df_line_graph$month == "Dec", ], aes(label = type)
            , hjust=0.9, vjust=-0.5) +
  scale_color_manual(values= c("#3182bd", "#bdbdbd")) +
  theme(legend.position=("none")
        , axis.line = element_line(color = "#bdbdbd")
        , axis.ticks = element_line(color = "#bdbdbd")
        , axis.text = element_text(color = "#bdbdbd")
  )
```

change the color of the axis titles to grey
```{r}
ggplot(df_line_graph, aes(x=month, y = volume, colour = type, group = type)) +
  geom_line(size=1) +
  theme_classic() +
  geom_text(data = df_line_graph[df_line_graph$month == "Dec", ], aes(label = type)
            , hjust=0.9, vjust=-0.5) +
  scale_color_manual(values= c("#3182bd", "#bdbdbd")) +
  theme(legend.position=("none")
        , axis.line = element_line(color = "#bdbdbd")
        , axis.ticks = element_line(color = "#bdbdbd")
        , axis.text = element_text(color = "#bdbdbd")
        , axis.title = element_text(color = "#bdbdbd")
  )
```